Popularity Adjusted Block Models are Generalized Random Dot Product Graphs

### 2022-01-11 referee-response.Rmd

6. p5. Definition 2. I do not understand the definition of the subset $\mathcal{X}$. 
Is it $x \in \mathcal{X}$ if for all $y \in \mathbb{R}^d$ we have $x^\top I_{p,q} y$ or is $\mathcal{X}$ a subset on $(\mathbb{R}^d)^2$? 
Please clearly state that $d = p + q$ (not stated before Definition 3).  
\textcolor{blue}{
We thank the referee for the suggestion. 
We clarify the definition of the subset as $\mathcal{X} = \{x, y \in \mathbb{R}^d : x^\top I_{p, q} y \in [0, 1]\}$ and add that $d = p + q$. 
}

7. p6. Remark 2. What are $\hat{Z}$, $V$, and $D$?  
\textcolor{blue}{
We thank the referee for the question. 
We now include in the remark the $D$ is the diagonal matrix of the $p$ most positive and $q$ most negative eigenvalues of $A$ (or $P$), $V$ is the matrix of corresponding eigenvectors, and $\hat{Z}$ is the embedding $V |D|^{1/2}$.
}

8. top of p7. $\tilde{P}$ instead of ${P}$.  
\textcolor{blue}{
We have corrected the typo. 
}

9. p7 line 40-46. I don't understand the interest of this paragraph.  
\textcolor{blue}{
We thank the referee for the question. 
The purpose of the additional material in the proof of Theorem 1 is to provide intuition for the arbitrary $K$ case. 
In particular, for the $K = 2$ case, we show that $\tilde{P} = X \Pi X^\top$ where $\Pi$ is a permutation matrix with 2 fixed points and 1 cycle of order 2, and the rows of $X$ consist of vectors that lie on two 2-dimensional orthogonal subspaces. 
The idea is that this generalizes to larger $K$ such that the rows of $X$ are vectors that lie on $K$ $K$-dimensional orthogonal subspaces, and $\Pi$ is a permutation matrix with $K$ fixed points and $K^2 - K = K (K - 1)$ cycles of order 2. 
Then the eigenvalues of $\Pi$ are broken up by the following: $K$ $1$'s corresponding to the $K$ fixed points and $K (K - 1)$ pairs of $+1$ and $-1$ corresponding to the $K (K - 1)$ cycles of order 2, 
resulting in $K (K + 1) / 2$ $+1$'s and $K (K - 1) / 2$ $-1$'s, i.e., the diagonal eigenvalue matrix of $\Pi$ is exactly $I_{K (K + 1) / 2, K (K - 1) / 2}$. 
This gives us the embedding dimension ($\mathbb{R}^{K^2}$ broken up into $K (K + 1) / 2$ ``positive'' and $K (K - 1) / 2$ ``negative'' dimensions according to the eigenvalues of $\Pi$) as well as the structure ($K$ $K$-dimensional subspaces as described by the rows of $X$). 
We have added additional details to the paper to address this. 
}

@@

10. p8, line 14-21. I don't see why the permutation given by $\Pi$ has $K$ fixed points.  
\textcolor{blue}{
We thank the referee for the comment. 
The purpose of Example 1 is to illustrate this explicitly. 
Comparing matrices $X$ and $Y$, we can see that they share three columns: the first, fifth, and ninth. 
The other columns are permuted. 
From Example 1 (as well as the proof of Theorem 1 for $K = 2$), it should be clear how this property holds for larger $K$. 
The proof of Theorem 1 for arbitrary $K$ should then confirm the intuition based on $K = 2$ and $3$. 
}

11. Section 3.1. I found it confusing that community detection algorithms are detailed when $P$ is assumed to be known and then methods for estimating $P$ are recalled. A sentence in the beginning of this section to expose what is presented could help the reader.  
\textcolor{blue}{
We thank the referee for the suggestion. 
The goal of estimation is not to estimate $P$ but to identify the community labels and estimate the popularity parameters. 
This is not immediately obvious even after observing $P$, observing $P$ does make estimation much easier. 
In the introduction to section 3 (before 3.1), we have added the following:
}  
\textcolor{blue}{
``In our methods, the data that are observed for estimation is the adjacency matrix, $A \sim \mathrm{PABM}(\{\lambda^{(k \ell)}\}_K, \rho_n)$, along with an assumed number of communities, $K$. 
To motivate some of the methods, we also consider community detection and parameter estimation in the case where we observe the edge probability matrix, $P$, noting that community memberships and popularity parameters are not immediately discernible from observing $P$.''
}

@@

12. p10, line 12. $c^{(i)}$ is not defined.  
\textcolor{blue}{
We thank the referee for pointing this out. 
$c^{(i)}$ is the $i^{th}$ entry of column vector $c$. 
This has been added to the paper.
}

13. p10. I didn't understand the exposition of the SSC algorithm.  
\textcolor{blue}{
We thank the referee for the question. 
We have added more detail to our description of the SSC algorithm. 
In particular, we talk a bit about the intuition/motivation behind the algorithm. 
This isn't added to the description, but $c$ and $c_i$ are column vectors of length $n$ and $c^{(i)}$ and $c_i^{(j)}$ are the $i^{th}$ and $j^{th}$ entries of vectors $c$ and $c_i$ respectively. 
}  
\textcolor{blue}{The description of SSC now reads:
}  
\textcolor{blue}{
``The SSC algorithm can be described as follows: 
Given \(X \in \mathbb{R}^{n \times d}\) with vectors
\(x_i^\top \in \mathbb{R}^d\) as rows of \(X\), the optimization problem
\(c_i = \arg\min_{c} \|c\|_1\) subject to \(x_i = X^\top c\) and
\(c^{(i)} = 0\), where $c^{(i)}$ is the $i^{th}$ entry of $c$, is solved for each \(i = [n]\). 
The solutions are collected collected into matrix
\(C = \bigl[ c_1 \mid \cdots \mid c_n \bigr]^\top\) to
construct an affinity matrix \(B = |C| + |C^\top|\). If each \(x_i\) lies
exactly on one of \(K\) subspaces, \(B\) describes an undirected graph
consisting of {\em at least} \(K\) disjoint subgraphs, i.e., \(B_{ij} = 0\) if \(x_i, x_j\) lie on different subspaces. 
The intuition here is that points that lie on the same subspace can be described as linear combinations of each other. 
Thus, for each $c_i$, $c_i^{(j)}$ is zero if $x_i$ and $x_j$ belong to different subspaces and may be nonzero if they do. 
If \(X\) instead represents points near \(K\) subspaces with some noise, 
then this property may only hold approximately and thus 
a final graph partitioning step may be required 
(e.g., edge thresholding or spectral clustering).''
}

@@

* This is clearly beyond the scope of the paper, but I wonder if there is a straightforward way to adapt the proposed methods to directed models or weighted graphs or any other types of graphs than binary undirected ones?  
\textcolor{blue}{
We thank the referee for the question. 
For directed graphs, the ASE provides a consistent estimator and thus should also work here \citep{doi:10.1080/01621459.2012.699795}. 
For weighted graphs, if $A_{ij}$ can be an estimator for $P_{ij}$ (e.g., if $A_{ij} \sim \mathrm{Poisson}(P_{ij})$), then all of the theory should transfer over to those models as well, 
and we believe that the theory can be adapted to other models where $g(A_{ij})$ is an estimator for $P_{ij}$. 
We hope to address these topics in future research. 
}

* In the simulation study, SSC-ASE works well for $K \geq 3$, but not for $K = 2$. 
The authors explain the bad performance by the use of gaussian mixtures in the last clustering step. 
However, I don't understand why this has an impact on the performance for $K = 2$, but not for $K \geq 3$.  
\textcolor{blue}{
We thank the referee for raising the concern. 
We are not entirely clear on why this happens for $K = 2$, but when comparing the normalized Laplacian eigenmaps of the affinity matrix $B$ for various $K$, we see that when $K > 2$, the points appear much closer to Gaussian mixtures, whereas with $K = 2$, there is some separation by community, but they do not appear as Gaussian mixtures, at least for the simulated data. 
One possible reason for this may be because we failed to find the right hyperparameter for SSC-ASE. 
}  
\textcolor{blue}{
Under ideal circumstances, we would be able to circumvent this issue altogether via the subspace detection property, but SDP does not necessarily guarantee a partitioning of the graph into exactly $K$ disjoint subgraphs but rather {\em at least} $K$ disjoint subgraphs, and in our simulations, we found that setting $\vartheta$ to the appropriate value to obtain SDP often results in greater than $K$ subgraphs, which is why we ended up choosing an ``incorrect'' $\vartheta$ for SSC and then performing the final clustering step.
}  
\textcolor{blue}{
We also note that in our simulations, while SSC-ASE and SSC-A in the $K = 2$ case and MM-Louvain for all $K$ output labels such that the number of mislabeled vertices increases with $n$, the proportion of mislabeled vertices still decreases.
}

### 2022-01-11 referee-response.tex

  \textcolor{blue}{
  We thank the referee for the question. 
  The purpose of the additional material in the proof of Theorem 1 is to provide intuition for the arbitrary $K$ case. 
  In particular, for the $K = 2$ case, we show that $\tilde{P} = X \Pi X^\top$ where $\Pi$ is a permutation matrix with 2 fixed points and 1 cycle of order 2, and the rows of $X$ consist of vectors that lie on two 2-dimensional orthogonal subspaces. 
  The idea is that this generalizes to larger $K$ such that the rows of $X$ are vectors that lie on $K$ $K$-dimensional orthogonal subspaces, and $\Pi$ is a permutation matrix with $K$ fixed points and $K^2 - K = K (K - 1)$ cycles of order 2. 
  Then the eigenvalues of $\Pi$ are broken up by the following: $K$ $1$'s corresponding to the $K$ fixed points and $K (K - 1)$ pairs of $+1$ and $-1$ corresponding to the $K (K - 1)$ cycles of order 2, 
  resulting in $K (K + 1) / 2$ $+1$'s and $K (K - 1) / 2$ $-1$'s, i.e., the diagonal eigenvalue matrix of $\Pi$ is exactly $I_{K (K + 1) / 2, K (K - 1) / 2}$. 

@@

  Section 3.1. I found it confusing that community detection algorithms
  are detailed when \(P\) is assumed to be known and then methods for
  estimating \(P\) are recalled. A sentence in the beginning of this
  section to expose what is presented could help the reader.\\
  The goal of estimation is not to estimate $P$ but to identify the community labels and estimate the popularity parameters. 
  This is not immediately obvious even after observing $P$, observing $P$ does make estimation much easier. 
  In the introduction to section 3 (before 3.1), we have added the following:
  ``In our methods, the data that are observed for estimation is the adjacency matrix, $A \sim \mathrm{PABM}(\{\lambda^{(k \ell)}\}_K, \rho_n)$, along with an assumed number of communities, $K$. 
  To motivate some of the methods, we also consider community detection and parameter estimation in the case where we observe the edge probability matrix, $P$, noting that community memberships and popularity parameters are not immediately discernible from observing $P$.''

@@

  One possible reason for this may be because we failed to find the right hyperparameter for SSC-ASE. 
  }\\
  \textcolor{blue}{
  Under ideal circumstances, we would be able to circumvent this issue altogether via the subspace detection property, but SDP does not necessarily guarantee a partitioning of the graph into exactly $K$ disjoint subgraphs but rather {\em at least} $K$ disjoint subgraphs, and in our simulations, we found that setting $\vartheta$ to the appropriate value to obtain SDP often results in greater than $K$ subgraphs, which is why we ended up choosing an ``incorrect'' $\vartheta$ for SSC and then performing the final clustering step.
  We also note that in our simulations, while SSC-ASE and SSC-A in the $K = 2$ case and MM-Louvain for all $K$ output labels such that the number of mislabeled vertices increases with $n$, the proportion of mislabeled vertices still decreases.

### 2022-01-11 summary_minh_edited.tex
finished responding to referees

It is nevertheless instructive to look at a few intermediate steps. 
More specifically, the product $U I_{3, 1} U^\top$ 
yields a permutation matrix $\Pi$ with fixed points at positions $1$ and $4$ 
and a cycle of order 2 swapping positions $2$ and $3$, i.e., 
$$\Pi = U I_{3, 1} U^\top = \begin{bmatrix} 1 & 0 & 0 & 0 \\
  0 & 0 & 1 & 0 \\
  0 & 1 & 0 & 0 \\
  0 & 0 & 0 & 1
\end{bmatrix}.$$
Furthermore, as $U$ is orthonormal and $I_{3, 1}$ is diagonal, 
$U I_{3, 1} U^\top$ is also an eigendecomposition of $\Pi$ where the fixed
points of $\Pi$ are mapped to the eigenvectors $e_1$ and $e_4$
while the cycles of order two are mapped to the eigenvectors  
$\tfrac{1}{\sqrt{2}}(e_{2} + e_3)$ and $\tfrac{1}{\sqrt{2}}(e_{2} -
e_3)$; here $e_i$ denote the $i^\mathrm{th}$ basis vector in $\mathbb{R}^{4}$. 
Thus, another way of decomposing the edge probability matrix is $\tilde{P} = X \Pi X^\top$ where the rows of $X$ describe 2 2-dimensional orthogonal subpsaces and $\Pi$ is a permutation matrix.

@@

In our methods, we assume that \(K\), the number of communities, is
known beforehand and does not require estimation.

@@

We previously stated in Theorem~\ref{theorem2} one possible set of latent positions that result in
the edge probability matrix of a PABM, namely
\(P = \tilde{\Pi} (XU) I_{p, q}
(XU)^\top \tilde{\Pi}^{\top}\) where $X$ is block diagonal and
$\tilde{\Pi}$ is a permutation matrix.  
Furthermore, the explicit form of \(XU\) represents points in \(\mathbb{R}^{K^2}\)
such that points within each community lie on \(K\)-dimensional
orthogonal subspaces, i.e. $\langle U^{\top} x_i, U^{\top} x_j \rangle = 0$ whenever $i$ and $j$ are in different communities. 
Thus ifwe have (or can estimate) \(XU\) directly, then both the community
detection and parameter identification problem are trivial because \(U\)
is orthonormal and fixed for each value of \(K\). 
However, direct
identification or estimation of \(XU\) is possibly difficult 
due to the non-identifiability of $XU$ (see Remark~\ref{rem:non_identifiable})
when we are given only $P$. 
More specifically, suppose we find a matrix $Y \in
\mathbb{R}^{n \times K^2}$
such that \(P = Y I_{p, q} Y^\top\). Then it is generally the case that
we  \( Y = \tilde{\Pi} XU Q \) for some indefinite orthogonal matrix
$Q \in \mathbb{O}(p,q)$. 
However since \(Q\) is not necessarily an
orthogonal matrix and hence, if $y_i$ denote the $i^{th}$ row of $Y$, 
then $\langle U^{\top} x_i, U^{\top} x_j \rangle
\neq \langle y_i, y_j \rangle$.
This prevents us from transferring the orthogonality property of
\(XU\) directly to $Y$.

@@

Theorem~\ref{theorem4} guarantees that for any $\epsilon > 0$, 
the number of edges of $\hat{B}$ between vertices of different communities 
that are larger than $\epsilon$ converges to zero with probability
converging to one as $n$ increases. 
We can always find an $\epsilon > 0$ such that $\hat{B}_{ij} > \epsilon$ 
with probability converging to one as $n$ increases. 
Thus, by using $\hat{B}$, we can 
perfectly recover all the latent community assignments $z_1, z_2,
\dots, z_n$, i.e., the number of mis-clustered vertices is zero
asymptotically almost surely. We note that Theorem~\ref{theorem4} is
stronger than existing results in the literature; in particular
Theorem~1 of \citet{307cbeb9b1be48299388437423d94bf1} (the paper that
originally introduces the PABM model) only guarantees that the {\em
  proportion} of mis-clustered vertices converges to $0$ as $n
\rightarrow \infty$. Furthermore Theorem~1
of \citet{307cbeb9b1be48299388437423d94bf1} also requires the sparsity
parameter $\rho_n$ to satisfies $n \rho_n^2 = \omega(\log^2{n})$ which
is a considerably stronger assumption than the assumption $n \rho_n =
\omega(\log^{4}{n})$ used in Theorem~\ref{theorem4}. Indeed, $n
\rho_n^2 = \omega(\log^{2}{n})$ implies $n \rho_n = \omega(n^{1/2})$. 
We emphasize that the assumption $n \rho_n = \omega(\log^{c}{n})$ for
some constant $c > 1$ is commonly used in the context of graph
estimation using spectral methods.

@@

From Fig.~\ref{fig:clust_err_ct_sim_imba} we once again see that the
number of mis-labeled vertices trending to $0$ for OSC. The
performance of SSC-ASE is comparable to that of OSC for $K > 2$ but
is worse when $K = 2$. Fig.~\ref{fig:lambda_est_p_imba} indicates that
the parameter estimation error also decays at rate $n^{-1/2}$ similar
to that in the balanced communities setting.

@@

Finally, one might well inquire what one gains and what one sacrifices
by assuming that a Bernoulli Graph is a PABM.  The GRDPG model offers
a plausible way to pursue this inquiry.  Absent a known latent
structure that can be exploited by specialized methods, the GRDPG-ASE
approach transforms the problem of network community detection to the
much-studied problem of clustering vectors in Euclidean space.
Communities of vertices are defined as clusters of latent vectors.
After ASE, a standard clustering algorithm, e.g., single linkage, is
used ot infer the communities.  In future research, we intend to use
such general algorithms as baselines and measure the efficiency of the
PABM algorithms (and other specialized algorithms) by studying how
much they improve on general algorithms when the specified latent
structure obtains.

>>>>

It is nevertheless instructive to look at a few intermediate steps. 
More specifically, the product $U I_{3, 1} U^\top$ 
yields a permutation matrix $\Pi$ with fixed points at positions $1$ and $4$ 
and a cycle of order 2 swapping positions $2$ and $3$, i.e., 
$$\Pi = U I_{3, 1} U^\top = \begin{bmatrix} 1 & 0 & 0 & 0 \\
  0 & 0 & 1 & 0 \\
  0 & 1 & 0 & 0 \\
  0 & 0 & 0 & 1
\end{bmatrix}.$$
Furthermore, as $U$ is orthonormal and $I_{3, 1}$ is diagonal, 
$U I_{3, 1} U^\top$ is also an eigendecomposition of $\Pi$ where the fixed
points of $\Pi$ are mapped to the eigenvectors $e_1$ and $e_4$
while the cycles of order two are mapped to the eigenvectors  
$\tfrac{1}{\sqrt{2}}(e_{2} + e_3)$ and $\tfrac{1}{\sqrt{2}}(e_{2} -
e_3)$; here $e_i$ denote the $i^\mathrm{th}$ basis vector in $\mathbb{R}^{4}$. 
Thus, another way of decomposing the edge probability matrix is $\tilde{P} = X \Pi X^\top$ where the rows of $X$ describe 2 2-dimensional orthogonal subspaces and $\Pi$ is a permutation matrix.

@@

In our methods, the data that are observed for estimation is the adjacency matrix, $A \sim \PABM(\{\lambda^{(k \ell)}\}_K, \rho_n)$, along with an assumed number of communities, $K$. 
To motivate some of the methods, we also consider community detection and parameter estimation in the case where we observe the edge probability matrix, $P$, noting that community memberships and popularity parameters are not immediately discernible from observing $P$.

@@

We previously stated in Theorem~\ref{theorem2} one possible set of latent positions that result in
the edge probability matrix of a PABM, namely
\(P = \tilde{\Pi} (XU) I_{p, q}
(XU)^\top \tilde{\Pi}^{\top}\) where $X$ is block diagonal and
$\tilde{\Pi}$ is a permutation matrix.  
Furthermore, the explicit form of \(XU\) represents points in \(\mathbb{R}^{K^2}\)
such that points within each community lie on \(K\)-dimensional
orthogonal subspaces, i.e. $\langle U^{\top} x_i, U^{\top} x_j \rangle = 0$ whenever $i$ and $j$ are in different communities. 
Thus if we have (or can estimate) \(XU\) directly, then both the community
detection and parameter identification problem are trivial because \(U\)
is orthonormal and fixed for each value of \(K\). 
However, direct
identification or estimation of \(XU\) is possibly difficult 
due to the non-identifiability of $XU$ (see Remark~\ref{rem:non_identifiable})
when we are given only $P$. 
More specifically, suppose we find a matrix $Y \in
\mathbb{R}^{n \times K^2}$
such that \(P = Y I_{p, q} Y^\top\). Then it is generally the case that
we  \( Y = \tilde{\Pi} XU Q \) for some indefinite orthogonal matrix
$Q \in \mathbb{O}(p,q)$. 
However since \(Q\) is not necessarily an
orthogonal matrix and hence, if $y_i$ denote the $i^{th}$ row of $Y$, 
then $\langle U^{\top} x_i, U^{\top} x_j \rangle
\neq \langle y_i, y_j \rangle$.
This prevents us from transferring the orthogonality property of
\(XU\) directly to $Y$.

@@

Theorem~\ref{theorem4} guarantees that for any $\epsilon > 0$, 
the number of edges of $\hat{B}$ between vertices of different communities 
that are larger than $\epsilon$ converges to zero with probability
converging to one as $n$ increases. 
We can always find an $\epsilon > 0$ such that $\hat{B}_{ij} > \epsilon$ 
with probability converging to one as $n$ increases. 
Thus, by using $\hat{B}$, we can 
perfectly recover all the latent community assignments $z_1, z_2,
\dots, z_n$, i.e., the number of misclustered vertices is zero
asymptotically almost surely. We note that Theorem~\ref{theorem4} is
stronger than existing results in the literature; in particular
Theorem~1 of \citet{307cbeb9b1be48299388437423d94bf1} (the paper that
originally introduces the PABM model) only guarantees that the {\em proportion} of misclustered vertices converges to $0$ as $n
\rightarrow \infty$. Furthermore Theorem~1
of \citet{307cbeb9b1be48299388437423d94bf1} also requires the sparsity
parameter $\rho_n$ to satisfies $n \rho_n^2 = \omega(\log^2{n})$ which
is a considerably stronger assumption than the assumption $n \rho_n =
\omega(\log^{4}{n})$ used in Theorem~\ref{theorem4}. Indeed, $n
\rho_n^2 = \omega(\log^{2}{n})$ implies $n \rho_n = \omega(n^{1/2})$. 
We emphasize that the assumption $n \rho_n = \omega(\log^{c}{n})$ for
some constant $c > 1$ is commonly used in the context of graph
estimation using spectral methods.

@@

From Fig.~\ref{fig:clust_err_ct_sim_imba} we once again see that the
number of mislabeled vertices trending to $0$ for OSC. The
performance of SSC-ASE is comparable to that of OSC for $K > 2$ but
is worse when $K = 2$. Fig.~\ref{fig:lambda_est_p_imba} indicates that
the parameter estimation error also decays at rate $n^{-1/2}$ similar
to that in the balanced communities setting.

@@

Finally, one might well inquire what one gains and what one sacrifices
by assuming that a Bernoulli Graph is a PABM.  The GRDPG model offers
a plausible way to pursue this inquiry.  Absent a known latent
structure that can be exploited by specialized methods, the GRDPG-ASE
approach transforms the problem of network community detection to the
much-studied problem of clustering vectors in Euclidean space.
Communities of vertices are defined as clusters of latent vectors.
After ASE, a standard clustering algorithm, e.g., single linkage, is
used to infer the communities.  In future research, we intend to use
such general algorithms as baselines and measure the efficiency of the
PABM algorithms (and other specialized algorithms) by studying how
much they improve on general algorithms when the specified latent
structure obtains.
